Implementing simplified INLA into `aghq`

### 2022-10-24 src/dev_sinla/sinla.Rmd
Loop i = 1:30 and plot the marginals

opt <- nlminb(
  start = h$par,
  objective = h$fn,
  gradient = h$gr,
  control = list(iter.max = its, trace = 0)
)
```

The values `opt$par` below correspond to $\hat \theta$:

```{r}
opt$par
```

We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt$par` above.
Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

```{r}
#' Notice that last.par contains opt$par as the fixed effects
#' It's the last.par because the last thing we've done with the template is the optimsation
#' This seems a pretty shakey way to go about things, but here we are...
mm <- h$env$last.par

@@

## $\tilde p(x_i \, | \, \theta, y)$

@@

Now we can loop over indices $i$ doing the required Laplace approximations.
In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.

```{r message=FALSE, warning=FALSE, results=FALSE}
objs <- list()
means <- list()
Qs <- list()

#' Starting parameters for TMB are the same for every index of the loop so it's more efficient to put them outside
param_i <- list(
  beta_prev = 0,
  phi_prev_i = 0,
  phi_prev_minus_i = rep(0, data$n - 1),
  log_sigma_phi_prev = 0
)

its <- 1000

for(i in 1:dat$n) {
  dat_i <- prepare_dat(dat, i)

@@

  opt <- nlminb(
    start = obj$par,
    objective = obj$fn,
    gradient = obj$gr,
    control = list(iter.max = its, trace = 0)
  )

  #' Parameters of the Gaussian approximation
  mm <- obj$env$last.par
  mean <- mm[obj$env$random]
  Q <- obj$env$spHess(mm, random = TRUE)

  objs[[i]] <- obj
  means[[i]] <- mean
  Qs[[i]] <- Q
}
```

Let's consider the first marginal, that is $\tilde p_\text{LA}(x_1 \, | \, \theta, y)$.
The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:

```{r message=FALSE, warning=FALSE}
i <- 1
obj <- objs[[i]]

mean <- means[[i]] #' The length is 30 - 1
length(mean)

Q <- Qs[[i]]
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
```

@@

## $\tilde p(x_i \, | \, y)$ {.tabset .tabset-fade}

@@

We're always going to need $\hat \theta$.
We have calculated it before, but we will get it fresh here (because of questionable programming practices whereby objects are overwritten):

```{r}
opt <- nlminb(
  start = h$par,
  objective = h$fn,
  gradient = h$gr,
  control = list(iter.max = its, trace = 0)
)

opt$par
```

Now, a basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
We can then create a dataframe where each row contains the values `opt$par` for $\theta$ and a grid value for $x_i$.
Each row can then be passed into `-obj$fn()` to calculate the log-posterior.
Note that it's crucial the arguments are in the right order when we pass them to `obj$fn()` -- this has caused headaches!

@@

df <- t(opt$par) %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(phi_prev_i = x_grid) %>%
  rowwise() %>%
  mutate(logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev)))
```

@@

### Safe exponential trapezoid rule {-}

@@

### `logSumExp` trapezoid rule {-}

@@

## Normalisation with `aghq`

@@

We could normalise $\tilde p_\text{LA}(\theta, y)$.
Note that sometimes we have been calling this $\tilde p_\text{LA}(\theta \, | \, y)$ because they are equivalent up to a constant of proportionality.
Define
$$
\tilde p_\text{AQ}(\theta \, | \, y) = \frac{\tilde p_\text{LA}(\theta, y)}{\tilde p_\text{AQ}(y)}
$$
or on a log-scale
$$
\log \tilde p_\text{AQ}(\theta \, | \, y) = \log \tilde p_\text{LA}(\theta, y) - \log \tilde p_\text{AQ}(y)
$$

@@

theta_names <- names(opt$par)
for(theta in theta_names) {
  param_i_fixed_theta[[theta]] <- opt$par[[theta]]
}

@@

@@

@@

Let $u_0$ be a zero column matrix.
I am guessing that this means $u_j$ is a $j$ column matrix.
It's not specified what dimension it is too, but since it's multiplied by $\mathcal{D}_{2j}$ presumably it must be of length $2j$.

@@

```{r}
D2j(3) %*% uj(3)
```

Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
Note that it hasn't been defined what $\Delta_j$ is yet.

@@

Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean):

```{r}
#' Conditional mean for the first grid point
Sigma <- solve(Q)
conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i])
```

Note: for this model the latent field is IID, so the conditional mean is the same as the mean.

@@

>>>>

opt_theta <- nlminb(
  start = h$par,
  objective = h$fn,
  gradient = h$gr,
  control = list(iter.max = its, trace = 0)
)
```

The values `opt_theta$par` below correspond to $\hat \theta$:

```{r}
opt_theta$par
```

We can also look at the parameters -- that is, mean vector and precision matrix -- of the Gaussian approximation at the mode `opt_theta$par` above.
Note that in `aghq`, what we would do is create a quadrature grid, then get the mean vector and precision matrix for each value of the quadrature grid over $\theta$.
By just sticking to the mode of $\theta$, we are effectively doing `aghq` with $k = 1$ in each dimension, i.e. one grid point, corresponding to empirical Bayes type of inference.

```{r}
#' Notice that last.par contains opt_theta$par as the fixed effects
#' It's the last.par because the last thing we've done with the template is the optimsation
#' This seems a pretty shakey way to go about things, but here we are...
mm <- h$env$last.par

@@

## For one index $i$

### $\tilde p(x_i \, | \, \theta, y)$

@@

Let's consider the first marginal, that is $\tilde p_\text{LA}(x_1 \, | \, \theta, y)$.
In the call to `MakeADFun` we set `random = "phi_prev_minus_i"` to integrate all of the random effects but that of the particular index.

```{r message=FALSE, warning=FALSE, results=FALSE}
i <- 1

param_i <- list(
  beta_prev = 0,
  phi_prev_i = 0,
  phi_prev_minus_i = rep(0, data$n - 1),
  log_sigma_phi_prev = 0
)

@@

opt <- nlminb(
  start = obj$par,
  objective = obj$fn,
  gradient = obj$gr,
  control = list(iter.max = its, trace = 0)
)
```

The mean and precision matrix for the Gaussian approximation $\tilde p_\text{G}(x_{-i} \, | \, x_i, \theta, y)$ are:

```{r}
mm <- obj$env$last.par
mean <- mm[obj$env$random]
length(mean) #' The length is 30 - 1

Q <- obj$env$spHess(mm, random = TRUE)
dim(Q) #' With dimension [(30 - 1) x (30 - 1)]
```

@@

### $\tilde p(x_i \, | \, y)$ {.tabset .tabset-fade}

@@

A basic way we could approach this is to define a grid of points that we would like to evaluate the posterior at.
We can then create a dataframe where each row contains the values `opt_theta$par` for $\theta$ and a grid value for $x_i$.
Each row can then be passed into `-obj$fn()` to calculate the log-posterior.
Note that it's crucial the arguments are in the right order when we pass them to `obj$fn()` -- this has caused headaches!

@@

df <- t(opt_theta$par) %>%
  as.data.frame() %>%
  slice(rep(1:n(), each = x_grid_length)) %>%
  mutate(phi_prev_i = x_grid) %>%
  rowwise() %>%
  mutate(logpost = -obj$fn(c(beta_prev, phi_prev_i, log_sigma_phi_prev)))
```

@@

#### Safe exponential trapezoid rule {-}

@@

#### `logSumExp` trapezoid rule {-}

@@

### Normalisation with `aghq`

@@

We could start by normalising $\tilde p_\text{LA}(\theta, y)$.
Note that sometimes we have been calling this $\tilde p_\text{LA}(\theta \, | \, y)$ because they are equivalent up to a constant of proportionality.
Define
$$
\tilde p_\text{AQ}(\theta \, | \, y) = \frac{\tilde p_\text{LA}(\theta, y)}{\tilde p_\text{AQ}(y)}
$$
or on a log-scale
$$
\log \tilde p_\text{AQ}(\theta \, | \, y) = \log \tilde p_\text{LA}(\theta, y) - \log \tilde p_\text{AQ}(y)
$$

@@

theta_names <- names(opt_theta$par)
for(theta in theta_names) {
  param_i_fixed_theta[[theta]] <- opt_theta$par[[theta]]
}

@@

### Comparison

@@

## Looping over index $i$

So we've done the approximation for the first index $i = 1$!
Now let's try computing $\tilde p(x_i \, | \, y)$ for every index $i$, using a for-loop to generate the required Laplace approximations, and tidying up and condensing the code used:

```{r}
xi_laplace_marginal <- function(i) {
  dat_i <- prepare_dat(dat, i)
  map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))
  param_i_fixed_theta <- list(beta_prev = 0, phi_prev_i = 0, phi_prev_minus_i = rep(0, data$n - 1), log_sigma_phi_prev = 0)
  for(theta in names(opt_theta$par)) param_i_fixed_theta[[theta]] <- opt_theta$par[[theta]]
  obj_fixed_theta <- MakeADFun(data = dat_i, parameters = param_i_fixed_theta, random = "phi_prev_minus_i", DLL = "model1index", silent = TRUE, map = map_fixed_theta)
  quad <- aghq::aghq(ff = obj_fixed_theta, k = 3, startingvalue = 0, control = aghq::default_control_tmb())
  pdf_and_cdf <- aghq::compute_pdf_and_cdf(quad)[[1]]
  return(pdf_and_cdf)
}

out <- lapply(1:dat$n, xi_laplace_marginal)

laplace_df <- bind_rows(out, .id = "index") %>%
  mutate(index = as.numeric(index))
```

Let's also get everything we need for the joint Gaussian marginal for each index:

```{r}
mm <- h$env$last.par
mean <- mm[h$env$random]
Q <- h$env$spHess(mm, random = TRUE)
Sigma <- solve(Q)

gaussian_df <- lapply(1:length(mean), function(i) {
  x <- seq(-5, 5, length.out = 1000)
  mean_i <- as.numeric(mean[i])
  sd_i <- as.numeric(sqrt(Sigma[i, i]))
  data.frame(index = i, x = x, pdf = dnorm(x, mean = mean_i, sd = sd_i))
}) %>%
  bind_rows()
```

```{r}
samples_tmbstan <- extract(mcmc, pars = "phi_prev") %>%
  as.data.frame() %>%
  pivot_longer(
    cols = everything(),
    names_to = "index",
    names_prefix = "phi_prev.",
    names_transform = as.integer
  )

plot <- ggplot(samples_tmbstan, aes(x = value)) +
  geom_histogram(aes(y = ..density..), alpha = 0.8, fill = cbpalette[7]) +
  facet_wrap(~index) +
  theme_minimal() +
  labs(x = "phi_prev", y = "Posterior PDF")

plot2 <- plot +
  geom_line(data = gaussian_df, aes(x = x, y = pdf), col = cbpalette[1])
```

```{r}
plot3 <- plot2 +
  geom_line(data = laplace_df, aes(x = theta, y = pdf), col = cbpalette[2])

plot3
```

@@

Let $u_0$ be a zero column matrix.
I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry.
It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$.

@@

The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$.
If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix.
Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$?
And $\text{length}(u_j) = 2j$ as discussed above?

Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
But it hasn't been defined what $\Delta_j$ is yet.

@@

Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean).
We start by computing the conditional mean for the first grid point:

```{r}
Sigma <- solve(Q)
conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i])
```

Note that for this model the latent field is IID, so the conditional mean is the same as the mean.
Now we use Newton's method to find $\mu(x_i)$.
This requires some function $f(\mu(x_i), \ldots) = 0$.

@@

```{r}
j <- 1
```
